Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-05-03 inst/doc/nonparametric-bayes.Rmd
draft for ERE

% Non-parametric approaches to optimal policy are more robust

```{r libraries, include=FALSE, cache=FALSE}
library(pdgControl)
library(nonparametricbayes)
library(reshape2)
library(ggplot2)
library(data.table)
library(tgp)
library(MCMCpack)
library(plyr)
library(knitcitations)
```

```{r plotting-options, echo=FALSE, include=FALSE, cache=TRUE}
## Dargh these are ignored if not placed here!?
options(xtable.type = 'latex', table.placement="H", xtable.print.comment=FALSE)
opts_chunk$set(cache.path = 'cache-pdf/', cache=TRUE)
opts_chunk$set(dev = 'Cairo_pdf', fig.width=5.5, fig.height=4)

# This stuff should be handled by knit script(?) (load ggplot2 first though)!
theme_set(theme_bw(base_size=10))
theme_update(panel.background = element_rect(fill = "transparent",colour = NA),
             plot.background = element_rect(fill = "transparent",colour = NA))
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r stateeq}
f <- RickerAllee
p <- c(2, 10, 5) 
K <- 10
allee <- 5
```

```{r sdp-pars, dependson="stateeq"}
sigma_g <- 0.05
sigma_m <- 0.0
z_g <- function() rlnorm(1, 0, sigma_g)
z_m <- function() 1+(2*runif(1, 0,  1)-1) * sigma_m
x_grid <- seq(0, 1.5 * K, length=101)
h_grid <- x_grid
profit <- function(x,h) pmin(x, h)
delta <- 0.01
OptTime <- 20  # stationarity with unstable models is tricky thing
reward <- 0
xT <- 0
seed_i <- 1
Xo <- K # observations start from
x0 <- Xo # simulation under policy starts from
Tobs <- 35
```

```{r obs, dependson="sdp-pars"}
obs <- sim_obs(Xo, z_g, f, p, Tobs=Tobs, nz=1, 
                harvest = sort(rep(seq(0, .5, length=7), 5)), seed = seed_i)
```

```{r mle, dependson="obs"}
alt <- par_est(obs,  init = c(r = p[1], 
                              K = mean(obs$x[obs$x>0]), 
                              s = sigma_g))
est <- par_est_allee(obs, f, p,  
                     init = c(r = p[1] + 1, 
                              K = p[2] + 2, 
                              C = p[3] + 2, 
                              s = sigma_g))
```

```{r gp-priors}
#inv gamma has mean b / (a - 1) (assuming a>1) and variance b ^ 2 / ((a - 2) * (a - 1) ^ 2) (assuming a>2)
s2.p <- c(5,5)  
tau2.p <- c(5,1)
d.p = c(10, 1/0.1, 10, 1/0.1)
nug.p = c(10, 1/0.1, 10, 1/0.1) # gamma mean
s2_prior <- function(x) dinvgamma(x, s2.p[1], s2.p[2])
tau2_prior <- function(x) dinvgamma(x, tau2.p[1], tau2.p[2])
d_prior <- function(x) dgamma(x, d.p[1], scale = d.p[2]) + dgamma(x, d.p[3], scale = d.p[4])
nug_prior <- function(x) dgamma(x, nug.p[1], scale = nug.p[2]) + dgamma(x, nug.p[3], scale = nug.p[4])
beta0_prior <- function(x, tau) dnorm(x, 0, tau)
beta = c(0)
priors <- list(s2 = s2_prior, tau2 = tau2_prior, beta0 = dnorm, nug = nug_prior, d = d_prior, ldetK = function(x) 0)
```

```{r gp, dependson=c("obs", "gp-priors")}
  gp <- bgp(X=obs$x, XX=x_grid, Z=obs$y, verb=0,
          meanfn="constant", bprior="b0", BTE=c(2000,16000,2),
          m0r1=FALSE, corr="exp", trace=TRUE, 
          beta = beta, s2.p = s2.p, d.p = d.p, nug.p = nug.p, tau2.p = tau2.p,
          s2.lam = "fixed", d.lam = "fixed", nug.lam = "fixed", tau2.lam = "fixed")      
```

```{r opt, dependson=c("gp", "mle")}
  OPT <- optimal_policy(gp, f, est$f, alt$f,
                        p, est$p, alt$p,
                        x_grid, h_grid, sigma_g, 
                        sigma_g, sigma_g, # est$sigma_g, alt$sigma_g, but those ests are poor
                        delta, xT, profit, reward, OptTime)
```

```{r sim, dependson="opt"}
dt <- simulate_opt(OPT, f, p, x_grid, h_grid, x0, z_g, profit)
```

@@

@@

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological interactions
involved, in practice real mechanisms are often unknown and the true
dynamics too complex to be captured by simple models [@refs]. While
simple mechanistic models can nevertheless provide important insights
into possible dynamics -- for instance, demonstrating that a vaccine
does not have to be 100% effective to eliminate the transmission of a
virus  -- such approaches are not well-suited for use in
forecasting outcomes of potential management options.  Non-parametric 
approaches offer a more flexible alternative that can both more accurately
reflect the data available while also representing greater uncertainty
in areas (of state-space) where data is lacking.  Ecological research
and management strategy should pay closer attention to the opportunities 
and challenges nonparametric modeling can offer.

@@

We argue that while simple mechanistically motivated models
may be best for the latter case [@Cuddington2013; @Geritz2012], such models 
can be not only inaccurate but misleading in quantitative decision making.
Nonparametric models can more realistically represent uncertainties outside 
of observed data while also better capturing the dynamics in the region
observed.  Advances in the theory and computational implementations of 
nonparametric methods make them ripe for such applications.  We use the classic
problem of optimal harvest of a marine fishery to illustrate how the 
nonparametric approach of Gaussian processes can be applied.

We imagine the manager seeks to maximize some known socio-economic value function, 
given their knowledge of the system. 
In particular, we will focus on the case in which the manager faces a series of decisions,
such as setting a harvest quota for a marine fishery each year based on observations
of the stock size the year before.

@@

The ecological dynamics of most systems of management interest are
typically both complex and unknown [@refs].  Despite this, quantitative
policy optimization in these systems has almost exclusively been based
on simple, mechanistically motivated parametric models.  There are
several reasons for this.  First, limited available data almost always
precludes any accurate estimate of more complicated models.  Second, the
computational demands of determining an optimal strategy in a sequential
decision-theory problem suffer from the "curse of dimensionality"
that makes it difficult to consider more complicated models, states,
or actions.  Unfortunately, these challenges also make it particularly
difficult to reflect the true uncertainty either in the parameters of
the given models (parametric uncertainty) or in the structure of the
models themselves (structural uncertainty).  Further unknowns such as
measurement uncertainty, parameter uncertainty, unobserved states,
knowledge of boundary conditions, etc. further compound the issue.
Though a hierarchical Bayesian approach provides a natural way to
address these from a statistical standpoint, formulating reasonable
parametric  descriptions of each form of uncertainty is a challenging
task in itself, let alone the computational difficulties of solving such
a system. @Cressie2009 provides a good account of the successes and
challenges of the approach.  Applying these approaches in the management
context of sequential decision making, in which forecasts must be obtained
over a range of possible actions and updated regularly as new information
arrives makes such an approach less feasible still.

An active literature and growing computational power over the past several decades have
only marginally improved this situation.  Parametric or structural uncertainty
can be introduced only by increasing the state space to reflect a distribution
of possible parameters [@refs] or the degree of belief in each of a set of possible models [@refs]
Frequently, the space of possible actions must then be reduced or the algorithms adjusted
by approximations to keep computations feasible.

<span style="color:#007FFF;"> ***This part is overstated and unnecessary***
Though machine learning approaches have begun to appear in the
ecological and conservation literature (e.g. in species distribution models),
including the Gaussian process based approach used here [@Munch2005],
overall machine learning remains an unfamiliar and somewhat untrusted approach for many ecologists.
One potential barrier to their adoption is the absence of a framework
for applying machine learning approaches to resource management problems. </span>

<span style="color:#007FFF;">***transition sentences!?***</span>

Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation, usually described
by a simple parametric model. Here we illustrate how a stochastic dynamic
programming algorithm can alternatively be driven by the predictions from
a Gaussian process -- a machine learning approximation to the state dynamics.

<span style="color:#007FFF;">***Okay, perhaps that's novel, but it's pretty trivial.  Isn't it
obvious to everyone that it's trivial?***</span>

<span style="color:#007FFF;">***paraphrase advantages of machine learning***</span>

Management goals / decision-theoretic approaches need accurate prediction
over relevant (short?) timescales more than reasonable long-term
mechanisms.  Machine-learning approaches may offer the benefit of the
hierarchical Bayesian approach without the practical and computational
limitations of their parametric kin.  Non-parametric models are flexible
enough to take maximum advantage of the data available, while being
appropriately ambiguous about the dynamics of a system in regions of
parameter space that have been poorly or never sampled.

@@

@@

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach [@refs] uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Though modifications of this approach mentioned in the 
introduction can permit additional sources of uncertainty such as measurement 
error in the stock assessment, implementation errors in the harvest policy, [@Clark1986; @Roughgarden1996; @Sethi2005]
uncertainty in parameters [@Mangel1985; @Schapaugh2013] or model structure [@Williams2001; @Athanassoglou2012],
we focus on the simplest example to illustrate the comparison, where only a 
single form of uncertainty (the growth shocks $Z_t$) is present in the underlying
model, making additional modes unnecessary. 
<span style="color:#007FFF;">***Add comment / apology on single species dynamics?*** </span>

We compare this approach to our alternative that uses a Gaussian Process (GP)
in place of the a given stock recruitment curve.  We briefly revisit
a description of the Gaussian Process approach that has appeared in 
recent literature before describing how we take the important step of 
adapting the GP to the decision theory problem that can still be solved
using SDP.

@@

The essence of the GP approach can be captured in the
following thought experiment: An exhaustive parametric approach to the
challenge of structural uncertainty might proceed by writing down all
possible functional forms for the underlying dynamical system with all
possible parameter values for each form, and then consider searching
over this huge space to select the most likely model and parameters;
or using a Bayesian approach, assign priors to each of these possible
models and infer the posterior distribution of possible models. The
GP approach can be thought of as a computationally
efficient approximation to this approach. GPs represent
a large class of models that can be though of as capturing or reasonably
approximating the set of models in this collection.  By modeling at the
level of the process, rather than the level of parametric equation,
we can more concisely capture the possible behavior of these curves.
In place of a parametric model of the dynamical system, the GP 
approach postulates a prior distribution of (n-dimensional)
curves that can be though of as approximations to a range of possible
(parametric) models that might describe the data. The GP allows us
to consider a set of possible curves simultaneously.  
<span style="color:#007FFF;"> ***Figure 1 include curves drawn from the posterior density?*** </span>

<span style="color:#007FFF;">***Do we need more specifics on Gaussian process as an approximation
to parametric models? Discussion of Gaussian process vs other machine
learning / forecasting approaches that have less clear statistical
foundations?  If so, does this all belong in the discussion?***</span>

@@

## Example in a bistable system

Concerns over the potential for tipping points in ecological dynamics
[@Scheffer2001] highlight the dangers of uncertainty in ecological
management and pose a substantial challenge to existing decision-theoretic
approaches [@Brozovic2011].  Because intervention is often too late 
after a tipping point has been crossed (but see @Hughes2013), management
is most often concerned with avoiding potentially catostrophic tipping
points before any data is available at or following a transition that
would more clearly reveal these regime shift dynamics [e.g. @Bestelmeyer2012].

<!-- Explain the connection to alternative stable states -->

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape parameter
$\sigma_g$.  We start with an example in which the parameters are $r =$
`r p[1]`, $K =$ `r p[2]`, $C =$ `r p[3]`, and  $\sigma_g =$ `r sigma_g`.

Both parametric and nonparametric approaches will require training
data on which to base their model of the process.  We generate the
training data under the model described in Eq 1 for `r Tobs` time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$.  For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the harvest.

@@

We estimate two parametric models from the data using a maximum likelihood
approach.  The first model is structurally identical to the true model
(Eq 1), differing only in that it's parameters are estimated from the
observed data rather than given.  The alternative model is the Ricker
model, which is structurally similar and commonly used in for such data.

The maximum likelihood estimates (MLE) will assume that the process noise $Z_t$
is log-normal, as in the simulation itself.  In reality the noise structure might 
differ from this assumption, making this a best-case scenario for the MLE approach.  
The GP must also make an assumption about the noise structure -- in this case we 
assume additive noise, though multiplicative noise would be more accurate.  This
will illustrate the flexibility of the GP in fitting different noise structures
even without knowledge of the correct assumptions.

The GP estimate is Bayesian approach which requires specification of
prior distributions for the hyper-parameters [see Appendix for details,
or @Rassmussen1996, @Munch2005 for background], as given in Table 1.
Then given the data, both the GP and the MLE parametric models can
be estimated as described above, as we implement in R in the code
provided in supplementary materials.

@@

Figure 1 shows the  inferred Gaussian Process compared to the true and the
MLE parametric models, predicting the relationship between observed stock
size (x-axis) to the stock size following recruitment the following year.
The gray shading shows the uncertainty estimated by the GP.  Unlike the
MLE estimates, the GP uncertainty changes depending on where the data is
available, allowing for a much greater level of uncertainty at low stock
sizes.  While it would be natural (and straight forward) to condition
the GP on passing through the origin (0,0) (see appendix), the estimate
shown here is based only on the observed data. The observed data from
which each model is estimated is also shown.  The observations come
from only a limited region of state space corresponding to unharvested
or weakly harvested system.  No observations occur at the theoretical
optimum (7.5 units, see Figure 2) or near the tipping point (5 units).

<!-- Would it be useful to show the uncertainty in the other models? -->

<!-- with (light grey) and without (darker grey) measurement error -->

```{r gp_plot, dependson=c("gp", "mle", "plotting-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance (grey shading)  is also shown.  The training data is also shown as black points. "}
tgp_dat <- 
    data.frame(  x = gp$XX[[1]], 
                 y = gp$ZZ.km, 
                 ymin = gp$ZZ.km - 2 * sqrt(gp$ZZ.ks2), 
                 ymax = gp$ZZ.km + 2 * sqrt(gp$ZZ.ks2),
                 ymin2 = gp$ZZ.mean - 2 * sqrt(gp$ZZ.vark), 
                 ymax2 = gp$ZZ.mean + 2 * sqrt(gp$ZZ.vark))
  true <- sapply(x_grid, f, 0, p)
  alt_mean <- sapply(x_grid, alt$f, 0, alt$p)
  est_mean <- sapply(x_grid, est$f, 0, est$p)
  models <- data.frame(x=x_grid, GP=tgp_dat$y, 
                       Parametric=est_mean, 
                       True=true, 
                       Structural=alt_mean)
  models <- melt(models, id="x")
  names(models) <- c("x", "method", "value")

ggplot(tgp_dat) + 
#  geom_ribbon(aes(x,y,ymin=ymin,ymax=ymax), fill="gray80") +    # measurement error + process error
  geom_ribbon(aes(x,y,ymin=ymin2,ymax=ymax2), fill="gray60") + # process error only
  geom_line(data=models, aes(x, value, col=method), alpha=0.8, lwd=1) + 
  geom_point(data=obs, aes(x,y)) + 
  xlab(expression(X[t])) + ylab(expression(X[t+1])) +
  scale_colour_manual(values=cbPalette) 
```

The resulting optimal management strategy based on each of the inferred
models is shown in Figure 2, against the optimal strategy given the
true underlying dynamics.  Policies are shown in terms of target
escapement, $S_t$.  Under models such as this a constant escapement
policy is expected to be optimal [@Reed1979], whereby population levels
below a certain size $S$ are unharvested, while above that size the harvest
strategy aims to return the population to $S$, resulting in the hockey-stick
shaped policies shown.  Only the GP policy deviates slightly from this.  The MLE parametric model
corresponding to parametric uncertainty alone (gold line labeled "Parametric", 
that is, using the same structural form as the underlying simulation) nevertheless overestimates 
the carrying capacity of the stock and therefore underestimates the harvest
rates, always waiting for the stock to rise further than is actually possible.
The Ricker model (green line, showing "Structural" inaccuracies) does not 
foresee the tipping point dynamics reflected in the concavity of the other three
models (Figure 1), and selects a much more aggressive harvesting strategy.

```{r policies_plot, dependson=c("opt", "plotting-options"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  "}
policies <- 
  melt(data.frame(stock=x_grid, 
                  GP = x_grid[OPT$gp_D], 
                  Parametric = x_grid[OPT$est_D],
                  True = x_grid[OPT$true_D],
                  Structural = x_grid[OPT$alt_D]),
       id="stock")
names(policies) <- c("stock", "method", "value")

ggplot(policies, aes(stock, stock - value, color=method)) +
    geom_line(alpha=0.7, lwd=1) + 
    xlab("stock size") + ylab("escapement")  +
    scale_colour_manual(values=cbPalette)
```

The consequences of managing 100 replicate realizations of the 
simulated fishery under each of the policies estimated is shown in Figure 3.  As expected
from the policy curves, the structurally correct model ("Parametric" uncertainty, gold) 
underharvests, leaving the stock to vary around it's unfished optimum.  The GP overharvests
the population slightly relative to the optimal value.  The structurally
incorrect Ricker model ("Structural", green line) overharvests the population passed
the tipping point consistently, resulting in the immediate crash of the stock and 
thus derives minimal profits.

***(may need to show harvest dynamics since
that is not clear from the figure! Also isn't general, sometimes does
optimally, sometimes over-fishes.  Perhaps need to show more examples.)
May need to show profits too?***

```{r sim_plot, dependson=c("sim", "plotting-options"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model."}
ggplot(dt) + 
    geom_line(aes(time, fishstock, 
                  group=interaction(reps, method), 
                  color=method), alpha=.1) +
    scale_colour_manual(values=c("GP" = cbPalette[1], "Parametric" = cbPalette[2], "True" = cbPalette[3], "Structural" = cbPalette[4]),  
                        guide = guide_legend(override.aes = list(alpha = 1)))
```

@@

While non-parametric Bayesian approaches will not always be preferable 
to simple mechanistic models, we highlight three aspects of the problem
consider here that make these methods particularly valuable.  These aspects
are common to many conservation decision making problems, which thus merit
greater use of non-parametric machine learning approaches that can best 
take advantage of them.

@@

<!--
<span style="color:#007FFF;"> ***wow, run-on tangent here?*** </span>
Predictive accuracy is not the goal of all modeling, as ecologists
have been observing for as long as they made models (perhaps none
more memorably than @Levins1969).  Mechanistic modeling is at its
most powerful not when it is used to directly forecast future states
but when it provides an understanding of how to approach a problem.
SIR-type models from the epidemiological literature are a good example.
While the simplest SIR models have little predictive power over the
outbreak intensity or timing at a particular location, they provide a
powerful understanding of the spread of an infection in terms of a single,
biologically meaningful parameter: $R_0$, the basic reproductive number.
From the model, it becomes clear that management need not vaccinate every
member of the population to stop the spread, but rather it suffices to
vaccinate a sufficient fraction of the population to reduce $R_0$ below 1.
-->

### 3. Time scale for new data

In the sequential decision making problem we considered, we are
presented with new data after each action.  The relevant timescale
for the prediction is thus not the long-term dynamics, which would be
wildly divergent, but the dynamics over this much shorter interval.
While ecologists may be hesitant to base continual management on a model
with obviously inaccurate long-term behavior, engineers tend to consider
the problem in frequency space and gravitate to the opposite position --
a good control model should prioritize high-frequency accuracy over low
frequency accuracy.  The differences in intuition may arise from the
timescales at which each profession can typically adjust their control
variables -- much faster for a control system of a chemical plant than
state policy for a natural resource.  Still, the lesson is clear: when
facing repeated management decisions over a short timescale, such as
setting annual harvests of a fishery, it may be more valuable to use a
machine learning algorithm that makes accurate year-ahead predictions
that capture some of the high-frequency fluctuations that appear only
as noise in a mechanistic model of the long-term population dynamics.

@@

<span style="color:#007FFF;"> Jargony and vague </span>
While we have highlighted certain generic of this problem that allow the
nonparametric approach to excel -- short timescales between successive
observations and actions, the accuracy in the appropriate region of state
space, the ability to express uncertainty outside the observed data --
there are equally several aspects for which the nonparametric Bayesian approach is
nevertheless at a relative disadvantage. In this simulated example, the underlying
dynamics are truly governed by a simple parametric model, allowing
the parametric approaches to be more accurate.  Similarly, because the
dynamics are  one-dimensional dynamics and lead to  stable nodes (rather
than other attractors such as limit-cycles resulting in oscillations),
the training data provides relatively limited information about the
dynamics.  For these reasons, we anticipate that in higher-dimensional
examples characteristic of ecosystem management problems that the machine
learning approach will prove even more valuable.

<span style="color:#007FFF;"> Data complexity. Perhaps too far out of scope... </span>
The nonparametric Bayesian approach is also better suited to complex and
disparate data.  Incorporating various sources of information into
mechanistic models can be an immensely difficult due to the increased
complexity involved.  Only thanks to tandem advances in increasing
computational power and hierarchical statistical methodology have we
been able to tackle such intuitively important complexity (and the
potentially new available data that accompanies it) such as spatial
distribution, heterogeneities of space, time, and individuals, to shift
to ecosystem-based approaches from single-species based approaches.
Without the need to formulate mechanisms, many modern machine learning
algorithms can leverage potential information from all available sources
of data directly.  The algorithms can recognize unanticipated or subtle
patterns in large data sets that enable more accurate predictions than
mechanistic models that are formulated at a more macroscopic level.

@@

<span style="color:#007FFF;"> 
***Possible additional discussion material?***

* Discuss constant escapement in model, in policies.
* Limitations of this comparison: Are the maximum-likelihood solutions
a straw man?
* Discussion of alternative related approaches: POMDP/MOMDP

>>>>

% Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

@@

Decision-theoretic methods often rely on simple parametric models of
ecological dynamics to compare the value of a potential sequence of
actions. Unfortunately, such simple models rarely capture the complexity
or uncertainty found in most real ecosystems. Non-parametric Bayesian
methods offer a promising statistical approach for predictive modeling
of ecological dynamics in regions of state space where the data is
adequate, while at the same time offering more flexible patterns with
greater uncertainty outside the observed data. This contrasts from simple
parametric models which provide relatively constant level of uncertainty
in regions with and without adequate data. The consequence of such
misplaced confidence outside the data can lead to highly undesirable
results that may be avoided with the more flexible non-parametric
Bayesian approach.

@@

Most management recommendations from the ecological literature are based
on (or at least motivated by) parametric models. Though in principle
these models reflect mechanistic underpinnings of biological interactions
involved, in practice real mechanisms are often unknown and the true
dynamics too complex to be captured by simple models. While
simple mechanistic models can nevertheless provide important insights
into possible dynamics -- for instance, demonstrating that a vaccine
does not have to be 100% effective to eliminate the transmission of a
virus  -- such approaches are not well-suited for use in
forecasting outcomes of potential management options.  Non-parametric 
approaches offer a more flexible alternative that can both more accurately
reflect the data available while also representing greater uncertainty
in areas (of state-space) where data is lacking.  Ecological research
and management strategy should pay closer attention to the opportunities 
and challenges nonparametric modeling can offer.

@@

Though simple mechanistically motivated models
may be best for the latter case [@Cuddington2013; @Geritz2012], such models 
can be not only inaccurate but misleading in quantitative decision making. This 
arises for two reasons:

1. We do not know what the correct parameter values are for the models.
1. We do not know what the correct models are for ecological systems.

The complexity of ecological interactions and a lack of data contribute
greatly to both of the problems. This concern is particularly acute
in case of ecological tipping points [@Barnosky2012; @Scheffer2001],
which arise from feedbacks common in ecological systems and can lead to a
sudden catastrophic transition to an undesirable state.  In general we do
not know where such tipping points are unless we have already observed
the transition, in which case it is frequently too late to respond.
Though there may exist early warning signals for certain kinds of these
transitions that are driven by slow changes [@Scheffer2009], we do not
know when, where, or how to apply them to the decision making context
more generally [@Boettiger2013].

In addition to facing the potentially grave consequences of such transitions,
if they do exist in a system of interest, this generally means that we lack
data in the region near and beyond a possible tipping point.  As a consequence,
our models are least accurate just where we need them to be most accurate.  
Unfortunately, parametric modeling approaches are not well-suited to this 
scenario, in which good data is avialble over only part of the relevant 
state-space.  Necessary assumptions about model structure in any parametric
approach can lead such models to perform very poorly, even when accounting
for parameter uncertainty.

Non-parametric models can more realistically represent uncertainties outside 
of observed data while also better capturing the dynamics in the region
observed.  Advances in the theory and computational implementations of 
nonparametric methods make them ripe for such applications.  We use the classic
problem of optimal harvest of a marine fishery to illustrate how the 
nonparametric approach of Gaussian processes can be applied.  We will 
compare Bayesian implementations of both nonparametric and parametric 
models, which best allow us to capture the uncertainty of model estimation
in either case and permits a more natural comparison between approaches.

### Terminology: "Non-parametric"

The term non-parametric to describe modeling approaches such as Gaussian
processes is a common source of confusion, notwithstanding the fact that
the model is still specified by parameters.  Some literature has introduced
the term "semi-parametric" in tacit acknowledgment of this, which no doubt
only contributes to the confusion.  The problem is further exacerbated by
the several meanings assigned to the term in statistics: (a) that the method 
does not assume some particular probability distribution (e.g. a non-parametric bootstrap)
or (b) that the method does not assume a fixed structure to the model.  Our 
use is that of (b), in contrast to the classical approaches that do.  The 
difference is clearest by way of example.

@@

The ecological dynamics of most systems of management interest are
typically both complex and unknown.  Despite this, quantitative
policy optimization in these systems has almost exclusively been based
on simple, mechanistically motivated parametric models.  There are
several reasons for this.  First, limited available data almost always
precludes any accurate estimate of more complicated models.  Second, the
computational demands of determining an optimal strategy in a sequential
decision-theory problem suffer from the "curse of dimensionality"
that makes it difficult to consider more complicated models, states,
or actions.  Unfortunately, these challenges also make it particularly
difficult to reflect the true uncertainty either in the parameters of
the given models (parametric uncertainty) or in the structure of the
models themselves (structural uncertainty).  Further unknowns such as
measurement uncertainty, parameter uncertainty, unobserved states,
knowledge of boundary conditions, etc. further compound the issue.
Though a hierarchical Bayesian approach provides a natural way to
address these from a statistical standpoint, formulating reasonable
parametric  descriptions of each form of uncertainty is a challenging
task in itself, let alone the computational difficulties of solving such
a system. @Cressie2009 provides a good account of the successes and
challenges of the approach.  Applying these approaches in the management
context of sequential decision making, in which forecasts must be obtained
over a range of possible actions and updated regularly as new information
arrives makes such an approach less feasible still.

<!-- cut?? -->

An active literature and growing computational power over the past several
decades have only marginally improved this situation.  Parametric or
structural uncertainty can be introduced only by increasing the state
space to reflect a distribution of possible parameters or
the degree of belief in each of a set of possible models Frequently,
the space of possible actions must then be reduced or the algorithms
adjusted by approximations to keep computations feasible.

<!-- cut?? -->

Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation, usually described
by a simple parametric model. Here we illustrate how a stochastic dynamic
programming algorithm can alternatively be driven by the predictions from
a Gaussian process.

@@

## Underlying Model

Concerns over the potential for tipping points in ecological dynamics
[@Scheffer2001] highlight the dangers of uncertainty in ecological
management and pose a substantial challenge to existing decision-theoretic
approaches [@Brozovic2011].  Because intervention is often too late 
after a tipping point has been crossed (but see @Hughes2013), management
is most often concerned with avoiding potentially catastrophic tipping
points before any data is available at or following a transition that
would more clearly reveal these regime shift dynamics [e.g. @Bestelmeyer2012].

Where $Z_t$ is multiplicative noise function with mean 1, representing
stochastic growth. We will consider log-normal noise with shape parameter
$\sigma_g$.  We start with an example in which the parameters are $r =2$,
$K=8$, $C=5$ and $\sigma_g = 0.1$.
<!-- ` r p[1]`, $K =$ ` r p[2]`, $C =$ ` r p[3]`, and  $\sigma_g =$ ` r sigma_g`. -->

Both parametric and nonparametric approaches will require training
data on which to base their model of the process.  We generate the
training data under the model described in Eq 1 for $T_{\textrm{obs}}=40$ time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$.  For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the harvest.

@@

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Uncertainty in the parameter estimates can 
be estimated and integrated over to determine the optimal policy under
under uncertainty [@Mangel1985; @Schapaugh2013]. Uncertainty in the model
structure itself can only be addressed in this approach by hypothesizing 
alternative model structures, and then performing some model choice or
model averaging  [@Williams2001; @Athanassoglou2012]. For simplicity 
of the comparison we consider only the case of a structurally correct
and structurally incorrect model, and estimate Bayesian posterior distributions
of the parameters.  We will also compare to a simpler maximum likelihood
estimate of the parameters that ignores the issue of parameter uncertainty.

We compare these parametric approaches to our alternative that uses a Gaussian Process (GP)
in place of a given stock recruitment curve.  We briefly revisit
a description of the Gaussian Process approach that has appeared in 
recent literature before describing how we take the important step of 
adapting the GP to the decision theory problem that can still be solved
using SDP.

@@

The essence of the GP approach can be captured in the
following thought experiment: An exhaustive parametric approach to the
challenge of structural uncertainty might proceed by writing down all
possible functional forms for the underlying dynamical system with all
possible parameter values for each form, and then consider searching
over this huge space to select the most likely model and parameters;
or using a Bayesian approach, assign priors to each of these possible
models and infer the posterior distribution of possible models. The
GP approach can be thought of as a computationally
efficient approximation to this approach. GPs represent
a large class of models that can be though of as capturing or reasonably
approximating the set of models in this collection.  By modeling at the
level of the process, rather than the level of parametric equation,
we can more concisely capture the possible behavior of these curves.
In place of a parametric model of the dynamical system, the GP 
approach postulates a prior distribution of (n-dimensional)
curves that can be though of as approximations to a range of possible
(parametric) models that might describe the data. The GP allows us
to consider a set of possible curves simultaneously.

@@

@@

We estimate posterior distributions for two parametric models: one using the structurally correct
model as given in Eq (1), which we refer to as the "Parametric Bayes" model,
and another using the familiar Ricker model, using a Gibbs sampler
as described (with source code) in the appendix).  In addition we 
estimate the parameters of the structurally correct model by maximum
likelihood.

The posterior distribution for the hyper-parameters of the Gaussian 
process model are estimated by Metropolis-Hastings algorithm, again with
details and code provided in the Appendix.  @Rasmussen2006 provides
an excellent general introduction to Gaussian Processes and @Munch2005 
first discusses their application in this context.

@@

![ Mean dynamics predicted by each model.](Figure1.pdf)

Figure 1 shows the mean inferred state space dynamics of each model
relative to the true model used to generate the data, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year.  All models except the MLE model
estimate a distribution around the means shown here, and all models 
estimate a level of process noise, which is independent of the state
value (x).  Note that in contrast to the  other models shown, the mean
Gaussian process corresponds to a distribution of curves - as indicated
by the gray band - which itself has a mean shown in black.  Note that 
this mean GP is thus more certain of the dynamics in the region where
data is available then where it is not.

While it would be natural (and straight forward) to condition
the GP on passing through the origin (0,0) (see appendix), the estimate
shown here is based only on the observed data. The observed data from
which each model is estimated is also shown.  The observations come
from only a limited region of state space corresponding to unharvested
or weakly harvested system.  No observations occur at the theoretical
optimum or near the tipping point.

![ Steady-state optimal policies by each model.](Figure2.pdf)

The resulting optimal management strategy based on each of the inferred
models is shown in Figure 2, against the optimal strategy given the
true underlying dynamics.  Policies are shown in terms of target
escapement, $S_t$.  Under models such as this a constant escapement
policy is expected to be optimal [@Reed1979], whereby population levels
below a certain size $S$ are unharvested, while above that size the harvest
strategy aims to return the population to $S$, resulting in the hockey-stick
shaped policies shown.

![ Managing the simulated system over 100 replicate instances under each policy](Figure3.pdf)

The consequences of managing 100 replicate realizations of the 
simulated fishery under each of the policies estimated is shown in Figure 3.  As expected
from the policy curves, the structurally correct model under-harvests,
leaving the stock to vary around it's un-fished optimum.  The structurally
incorrect Ricker model over-harvests the population passed
the tipping point consistently, resulting in the immediate crash of the stock and 
thus derives minimal profits.

@@

While non-parametric Bayesian approaches will not always be preferable 
to simple mechanistic models, we highlight three aspects of the problem
consider here that make these methods particularly valuable.  These aspects
are common to many conservation decision making problems, which thus merit
greater use of non-parametric approaches that can best take advantage of them.

@@

@@

In this simulated example, the underlying
dynamics are truly governed by a simple parametric model, allowing
the parametric approaches to be more accurate.  Similarly, because the
dynamics are  one-dimensional dynamics and lead to  stable nodes (rather
than other attractors such as limit-cycles resulting in oscillations),
the training data provides relatively limited information about the
dynamics.  For these reasons, we anticipate that in higher-dimensional
examples characteristic of ecosystem management problems that the machine
learning approach will prove even more valuable.

<span style="color:#007FFF;"> Data complexity. Perhaps too far out of scope... </span>
The nonparametric Bayesian approach is also better suited to complex and
disparate data.  Incorporating various sources of information into
mechanistic models can be an immensely difficult due to the increased
complexity involved.

### Online learning

In our treatment here we have ignored the possibility of learning during the 
management phase, in which the additional observations of the stock size could
potentially improve parameter estimates.  While we intend to address this 
possibility in future work in the context of these non-parametric models,
we have not addressed it here for pedagogical reasons. In the context presented
here, it is clear that the differences in performance arise from differences
in the uncertainty inherent in the model formulations, rather than from 
differing abilities to learn.  Because we consider a threshold system, 
online learning would not change this generic feature of a lack of data in a
certain range of the state space which is better captured by the Gaussian process.

@@

Appendix
========

The appendices have not yet been assembled.  Meanwhile, code to repeat the analyses, along with a complete log of all research conducted on this project, can be found at: [https://github.com/cboettig/nonparametric-bayes](https://github.com/cboettig/nonparametric-bayes/)

## Markov Chain Monte Carlo Analysis

![Traces of the GP hyper-parameters (after transforming back from log space)](gp_traces_densities1.pdf)
